---
title: "Covid NP Microbiome Baseline Statistics"
author: "Youngapelian, M.J."
date: "3/3/2022"
output: html_document
---

# load libraries
```{r include = FALSE}
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(ggsci)
library(data.table)
library(dataRetrieval)
library(readr)
library(stats)
library(psych)
```

# load metadata files from original and final filtered metadata
```{r include = FALSE}
full_metadata <- read_delim("/Users/mykajaapyoungapelian/covid-microbiome/metadataFiles/20220203_COVID_metadata.txt", "\t", escape_double = FALSE, trim_ws = TRUE)

filtered_metadata <- read_delim("/Users/mykajaapyoungapelian/covid-microbiome/metadataFiles/finalMetadata/physeq_gen_final_mapping.txt", "\t", escape_double = FALSE, trim_ws = TRUE)
```

```{r}
# create variables as numeric
d <- data.table(full_metadata)

# changing variables to numeric variables
d$gender <- as.numeric(d$gender)
d$death <- as.numeric(d$death)
d$severity <- as.numeric(d$severity)
d$smoking_status <- as.numeric(d$smoking_status)
d$hosp_transfer <- as.numeric(d$hosp_transfer)
d$icu_admit_UMCU <- as.numeric(d$icu_admit_UMCU)
d$CCI <- as.numeric(d$CCI)
d$fever_ceoccur_v2 <- as.numeric(d$fever_ceoccur_v2)
d$fatigue <- as.numeric(d$fatigue)
d$respiratory <- as.numeric(d$respiratory)
d$oxygen_cmoccur <- as.numeric(d$oxygen_cmoccur)
d$pre_antibiotic_use_2 <- as.numeric(d$pre_antibiotic_use_2)
d$pre_antibiotic_category <- as.numeric(d$pre_antibiotic_category)
d$antibiotic_cat_1 <- as.numeric(d$antibiotic_cat_1)
d$antibiotic_cat_2 <- as.numeric(d$antibiotic_cat_2)
d$antibiotic_cat_3 <- as.numeric(d$antibiotic_cat_3)
d$Ct <- as.numeric(d$Ct)
d$`ng/ul` <- as.numeric(d$`ng/ul`)
d$antibiotics_during_admit <- as.numeric(d$antibiotics_during_admit)
d$switched_antibiotics_at_umcu <- as.numeric(d$switched_antibiotics_at_umcu)
#summary(d)
```

# subset samples on covid positive / negative: age statistics
```{r}
covid_pos <- d[covid %in% c(1)]
round(mean(covid_pos$age), 1) # mean age covid pos 63.3
round(sd(covid_pos$age), 1) # sd age covid pos 14.2 
length(covid_pos$age) # 189
range(covid_pos$age) # 25 - 91
#hist(covid_pos$age)
#qqnorm(covid_pos$age) 
#qqline(covid_pos$age)  

covid_neg <- d[covid %in% c(0)]
round(mean(covid_neg$age), 1) # mean age covid neg 52.3
round(sd(covid_neg$age), 1) # sd age covid neg 18.5
length(covid_neg$age) # 71
range(covid_neg$age) # 21 - 90
#hist(covid_neg$age)
#qqplot(covid_neg$age)
```

# severity baseline statistics
```{r include = FALSE}
covid_pos$severity <- factor(covid_pos$severity, levels = c(1, 2, 3, 4))
boxplot(covid_pos$severity)

# summary statistics on age by severity 
by(covid_pos$age, covid_pos$severity, describe)

# severity group statistics across age (mean, sd, and range)
tapply(covid_pos$age, covid_pos$severity, mean)
tapply(covid_pos$age, covid_pos$severity, sd)
tapply(covid_pos$age, covid_pos$severity, range)
#boxplot(covid_pos$age ~ covid_pos$severity)

# severity counts per group
d[, .(x = sum(severity == 0))] #71 [27.3 %]
d[, .(x = sum(severity == 1))] #34 [18.0 %]
d[, .(x = sum(severity == 2))] #39 [20.6 %]
d[, .(x = sum(severity == 3))] #79 [41.8 %]
d[, .(x = sum(severity == 4))] #37 [19.6 %]
round(100*(34/(71+ 34+39+79+37)), 1) #  calculate percentage pos/neg per severity
round(100*(34/(34+39+79+37)), 1) # calculate percentage positive per severity 
```

# anova on age across severity levels 
```{r}
age_severity <- covid_pos[, .(age), by = severity]

age_severity_model <- aov(age_severity$age ~ factor(age_severity$severity))
summary(age_severity_model) #H0: Same mean; #H1: Different means 
# p = 0.000961, at least one of the severity levels differs among age

# complete tukey's post-hoc test for each comparison
TukeyHSD(age_severity_model, conf.level = .95)
# difference between mild-death (padj = 0.0066685) and moderate-death (padj = 0.0062876)
```

# t test on age across positive and negative samples 
```{r}
t.test(covid_pos$age, covid_neg$age)
 #H0: Same mean; #H1: Different means 
# p = 1.71e-05, the mean age between pos/neg is different
#95 percent confidence interval: [6.150937; 15.797196]
# sample estimates: mean of x: 63.26984; mean of y 52.29577
```

# fisher exact test on gender contingency table across severity 
```{r}
# severity count by gender 
covid_pos[, .(x = sum(severity == 1)), by = gender]
covid_pos[, .(x = sum(severity == 2)), by = gender]
covid_pos[, .(x = sum(severity == 3)), by = gender]
covid_pos[, .(x = sum(severity == 4)), by = gender]

# gender proportion tables across severity 
table_gender_severity <- table(covid_pos$gender, covid_pos$severity, useNA = "ifany")
margin.table(table_gender_severity, 1) # calculate total genders
margin.table(table_gender_severity, 2) # calculate all genders across severity levels 
prop.table(table_gender_severity,1) # calculates proportions of variable across rows 
prop.table(table_gender_severity,2) # calculates proportions of variable across columns 

# fisher exact test
fisher.test(table(covid_pos$gender, covid_pos$severity)) # without NA p = 0.0885
fisher.test(table_gender_severity) # with NA: p-value = 0.0004022
```

# fisher exact test on gender contingency table across pos/neg 
```{r}
# pos/neg count by gender 
covid_pos[, .(x = sum(covid == 1)), by = gender]
covid_neg[, .(x = sum(covid == 0)), by = gender]

# gender proportion tables across pos/neg 
table_gender_pos_neg <- table(d$gender, d$covid, useNA = "ifany")
margin.table(table_gender_pos_neg, 1) # calculate total genders
margin.table(table_gender_pos_neg, 2) # calculate all genders across pos/neg levels 
prop.table(table_gender_pos_neg,1) # calculates proportions of variable across rows 
prop.table(table_gender_pos_neg,2) # calculates proportions of variable across columns 

# fisher exact test
fisher.test(table(d$gender, d$covid)) # without NA p = 0.2234
fisher.test(table_gender_pos_neg) # with NA: p-value = 0.005
```

# smoking status baseline statistics by severity
```{r}
# 0 - no, 1 - yes, 2 - former, 3 - unknown, NA - Unknown 
table_smoking_severity_NA <- table(covid_pos$smoking_status, covid_pos$severity, useNA = "always")
table_smoking_severity <- table(covid_pos$smoking_status, covid_pos$severity)

margin.table(table_smoking_severity_NA, 1) # calculate total smoking_status
round(100*(margin.table(table_smoking_severity_NA, 1)/189), 1)

margin.table(table_smoking_severity_NA, 2) # calculate all smoking_status across severity levels

prop.table(table_smoking_severity_NA,1) # calculates proportions of variable across rows 
prop.table(table_smoking_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_smoking_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_smoking_severity, simulate.p.value = TRUE) # without NA  and simulated pval p = 0.0896
```

# mean charleson comorbidity index baseline statistics by severity
```{r}
# summary statistics on cci by severity 
cci_severity <- covid_pos[, .(CCI), by = severity]
cci_severity[!is.na(CCI), .N, by = severity]
cci_severity[!is.na(CCI), .(x_bar = mean(CCI)), by = severity]
cci_severity[!is.na(CCI), .(x_sd = SD(CCI)), by = severity]
cci_severity[!is.na(CCI), .(x_range = range(CCI)), by = severity]
# overall mean, sd, and range 
cci_severity[!is.na(CCI), .(x_bar = mean(CCI), x_sd = sd(CCI), x_range = range(CCI))]

# anova of CCI amongst severity groups 
cci_severity_model <- aov(cci_severity$CCI ~ factor(cci_severity$severity))
summary(cci_severity_model) #H0: Same mean; #H1: different means 
# p = 0.078, we can assume similar means 
```

# hospital transfered baseline statistics by severity 
```{r}
# 0 - no; # 1 - yes, NA 
table_transfer_severity_NA <- table(covid_pos$hosp_transfer, covid_pos$severity, useNA = "always")
table_transfer_severity <- table(covid_pos$hosp_transfer, covid_pos$severity)


margin.table(table_transfer_severity_NA, 1) # calculate total hosp_transfer
round(100*(margin.table(table_transfer_severity_NA, 1)/189), 1)

#margin.table(table_transfer_severity_NA, 2) # calculate total group n across severity levels
table_transfer_severity_NA # calculate each group n across severity 
prop.table(table_transfer_severity_NA,1) # calculates proportions of variable across rows 
prop.table(table_transfer_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_transfer_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_transfer_severity) # without NA pval p = 8.157e-06 
```

# admitted to icu baseline statistics by severity 
```{r}
# 0 - no; # 1 - yes, NA 
table_icu_severity_NA <- table(covid_pos$icu_admit_UMCU, covid_pos$severity, useNA = "always")
table_icu_severity <- table(covid_pos$icu_admit_UMCU, covid_pos$severity)


margin.table(table_icu_severity_NA, 1) # calculate total icu_admit_UMCU
round(100*(margin.table(table_icu_severity_NA, 1)/189), 1)

#margin.table(table_icu_severity_NA, 2) # calculate total group n across severity levels
table_icu_severity_NA # calculate each group n across severity 
#prop.table(table_icu_severity_NA,1) # calculates proportions of variable across rows 
prop.table(table_icu_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_icu_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_icu_severity) # without NA pval p = 2.62e-09 
```

# pre-antibiotic use baseline statistics 
```{r}
# 1 - yes, NA - unknown
table_pre_antibiotic_severity_NA <- table(covid_pos$pre_antibiotic_use_2, covid_pos$severity, useNA = "always")
table_pre_antibiotic_severity <- table(covid_pos$pre_antibiotic_use_2, covid_pos$severity, useNA = "no")


margin.table(table_pre_antibiotic_severity_NA, 1) # calculate total pre_antibiotic_use_2
round(100*(margin.table(table_pre_antibiotic_severity_NA, 1)/189), 1)

#margin.table(table_pre_antibiotic_severity_NA, 2) # calculate total group n across severity levels
table_pre_antibiotic_severity_NA # calculate each group n across severity 
#prop.table(table_pre_antibiotic_severity_NA,1) # calculates proportions of variable across rows 
#prop.table(table_pre_antibiotic_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_pre_antibiotic_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_pre_antibiotic_severity) # with NA pval p = 0.0001291
```

# mean bacterial dna baseline statistics 
```{r}
# summary statistics on bacterial dna by severity 
bacterial_dna_severity <- covid_pos[, .(Ct), by = severity]
bacterial_dna_severity[!is.na(Ct), .N, by = severity]

# overall mean, sd, and range
bacterial_dna_severity[!is.na(Ct), .(x_bar = round(mean(Ct),1), x_sd = round(sd(Ct), 1), x_range = round(range(Ct), 1))]

# overall mean, sd, and range per group
bacterial_dna_severity[!is.na(Ct), .(x_bar = round(mean(Ct),1), x_sd = round(sd(Ct), 1), x_range = round(range(Ct), 1)), by = severity]

# anova of Ct amongst severity groups 
bacterial_dna_severity_model <- aov(bacterial_dna_severity$Ct ~ factor(bacterial_dna_severity$severity))
summary(bacterial_dna_severity_model) #H0: Same mean; #H1: different means 
# p = 0.681, we can assume similar means 
```

# fever at admission baseline statistics
```{r}
# 0 - no, 1 - yes, NA - unknown
table_fever_severity_NA <- table(covid_pos$fever_ceoccur_v2, covid_pos$severity, useNA = "always")
table_fever_severity <- table(covid_pos$fever_ceoccur_v2, covid_pos$severity, useNA = "no")

# calculate total fever_ceoccur_v2
margin.table(table_fever_severity_NA, 1) 
round(100*(margin.table(table_fever_severity_NA, 1)/189), 1)

#margin.table(table_fever_severity_NA, 2) # calculate total group n across severity levels
table_fever_severity_NA # calculate each group n across severity 
#prop.table(table_fever_severity_NA,1) # calculates proportions of variable across rows 
#prop.table(table_fever_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_fever_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_fever_severity) # pval p = 0.3624
```

# oxygen at admisison basleine statistics 
```{r}
# 0 - no, 1 - yes, NA - unknown
table_oxygen_severity_NA <- table(covid_pos$oxygen_cmoccur, covid_pos$severity, useNA = "always")
table_oxygen_severity <- table(covid_pos$oxygen_cmoccur, covid_pos$severity, useNA = "no")


margin.table(table_oxygen_severity_NA, 1) # calculate total oxygen_cmoccur
round(100*(margin.table(table_oxygen_severity_NA, 1)/189), 1)

# margin.table(table_oxygen_severity_NA, 2) # calculate total group n across severity levels
table_oxygen_severity_NA # calculate each group n across severity 
# prop.table(table_oxygen_severity_NA,1) # calculates proportions of variable across rows 
# prop.table(table_oxygen_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_oxygen_severity_NA,2)), 1)

# fisher exact test 
fisher.test(table_oxygen_severity) # pval p = 0.6857
```

# fatigue at admission baseline statistics 
```{r}
# 0 - no, 1 - yes, NA - unknown
table_fatigue_severity_NA <- table(covid_pos$fatigue, covid_pos$severity, useNA = "always")
table_fatigue_severity <- table(covid_pos$fatigue, covid_pos$severity, useNA = "no")


margin.table(table_fatigue_severity_NA, 1) # calculate total fatigue
round(100*(margin.table(table_fatigue_severity_NA, 1)/189), 1)

#margin.table(table_fatigue_severity_NA, 2) # calculate total group n across severity levels
table_fatigue_severity_NA # calculate each group n across severity 
#prop.table(table_fatigue_severity_NA,1) # calculates proportions of variable across rows 
#prop.table(table_fatigue_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_fatigue_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_fatigue_severity) # with pval p = 0.2104
```


# started antibiotics during UMCU stay baseline statistics 
```{r}
# % change all dates to as.Date format
# covid_pos$antibiotic_1_start_date <- as.Date(covid_pos$antibiotic_1_start_date)
# covid_pos$antibiotic_2_start_date <- as.Date(covid_pos$antibiotic_2_start_date)
# covid_pos$antibiotic_3_start_date <- as.Date(covid_pos$antibiotic_3_start_date)
# covid_pos$presentation_UMCU <- as.Date(covid_pos$presentation_UMCU)

# if antibiotic start date was before presentation date then they did not start antibiotics at UMCU
antibitoics_no_start_umcu <- covid_pos[presentation_UMCU < antibiotic_1_start_date | presentation_UMCU < antibiotic_2_start_date | presentation_UMCU < antibiotic_3_start_date]

# never took antibiotics 
antibiotic_never_umcu <- covid_pos[is.na(pre_antibiotic_use_2) & is.na(antibiotic_1_start_date) & is.na(antibiotic_2_start_date) & is.na(antibiotic_3_start_date)]

# select all rows where all medication start dates are >= to umcu admit date
antibiotic_start_umcu_1 <- covid_pos[presentation_UMCU <= antibiotic_1_start_date | presentation_UMCU <= antibiotic_2_start_date | presentation_UMCU <= antibiotic_3_start_date, ab_start_umcu := "yes"]
# 189 took antibiotics at UMCU






```


# switched to other antibiotics during umcu stay baseline statistics
```{r}
antibiotic_switch <- 
```

# proportion table of antibiotic categories 
```{r}
antibiotic_categories <- d[, .(pre_antibiotic_category, antibiotic_cat_1, antibiotic_cat_2, antibiotic_cat_3)]

antibiotic_counts <- table(c(antibiotic_categories$pre_antibiotic_category, antibiotic_categories$antibiotic_cat_1, antibiotic_categories$antibiotic_cat_2, antibiotic_categories$antibiotic_cat_3),  useNA = "no")

round(100* prop.table(antibiotic_counts), 1)
```


#######################################################################
# baseline table statistics following exclusion criteria
```{r}
# create variables as numeric
d <- data.table(filtered_metadata)

# changing variables to numeric variables
d$gender <- as.numeric(d$gender)
d$death <- as.numeric(d$death)
#d$severity <- as.numeric(d$severity)
d$smoking_status <- as.numeric(d$smoking_status)
d$hosp_transfer <- as.numeric(d$hosp_transfer)
d$icu_admit_UMCU <- as.numeric(d$icu_admit_UMCU)
d$CCI <- as.numeric(d$CCI)
d$fever_ceoccur_v2 <- as.numeric(d$fever_ceoccur_v2)
d$fatigue <- as.numeric(d$fatigue)
d$respiratory <- as.numeric(d$respiratory)
d$oxygen_cmoccur <- as.numeric(d$oxygen_cmoccur)
d$pre_antibiotic_use_2 <- as.numeric(d$pre_antibiotic_use_2)
d$pre_antibiotic_category <- as.numeric(d$pre_antibiotic_category)
d$antibiotic_cat_1 <- as.numeric(d$antibiotic_cat_1)
d$antibiotic_cat_2 <- as.numeric(d$antibiotic_cat_2)
d$antibiotic_cat_3 <- as.numeric(d$antibiotic_cat_3)
d$Ct <- as.numeric(d$Ct)
d$`ng/ul` <- as.numeric(d$`ng/ul`)
d$antibiotics_during_admit <- as.numeric(d$antibiotics_during_admit)
d$switched_antibiotics_at_umcu <- as.numeric(d$switched_antibiotics_at_umcu)
#summary(d)
```

# subset samples on covid positive / negative: age statistics
```{r}
covid_pos <- d[covid %in% c("positive")]
round(mean(covid_pos$age), 1) # mean age covid pos 63.3
round(sd(covid_pos$age), 1) # sd age covid pos 14.2 
length(covid_pos$age) # 189
range(covid_pos$age) # 25 - 91
#hist(covid_pos$age)
#qqnorm(covid_pos$age) 
#qqline(covid_pos$age)  

covid_neg <- d[covid %in% c("negative")]
round(mean(covid_neg$age), 1) # mean age covid neg 52.3
round(sd(covid_neg$age), 1) # sd age covid neg 18.5
length(covid_neg$age) # 71
range(covid_neg$age) # 21 - 90
#hist(covid_neg$age)
#qqplot(covid_neg$age)
```


# severity baseline statistics
```{r include = FALSE}
covid_pos$severity <- factor(covid_pos$severity, levels = c("mild", "moderate", "severe", "death"))
#boxplot(covid_pos$severity)

# summary statistics on age by severity 
by(covid_pos$age, covid_pos$severity, describe)

# severity group statistics across age (mean, sd, and range)
round(tapply(covid_pos$age, covid_pos$severity, mean), 1)
round(tapply(covid_pos$age, covid_pos$severity, sd), 1)
tapply(covid_pos$age, covid_pos$severity, range)
#boxplot(covid_pos$age ~ covid_pos$severity)

# severity counts per group
d[, .(x = sum(covid == "positive")), by = severity]
round(100*(25/(91)), 1) # calculate percentage positive per severity 
```

# anova on age across severity levels 
```{r}
age_severity <- covid_pos[, .(age), by = severity]

age_severity_model <- aov(age_severity$age ~ factor(age_severity$severity))
summary(age_severity_model) #H0: Same mean; #H1: Different means 
# p = 0.0473, at least one of the severity levels differs among age

# complete tukey's post-hoc test for each comparison
TukeyHSD(age_severity_model, conf.level = .95)
# difference between mild-death (padj = 0.0493960)
```

# t test on age across positive and negative samples for selected
```{r}
t.test(covid_pos$age, covid_neg$age)
 #H0: Same mean; #H1: Different means 
# p = 0.0095, the mean age of pos/neg are different
#95 percent confidence interval: [1.952942; 13.684421]
# sample estimates: mean of x: 61.31868; mean of y 53.50000
```
# fisher exact test on gender contingency table across pos/neg 
```{r}
# pos/neg count by gender 
covid_pos[, .(x = sum(covid == "positive")), by = gender]
covid_neg[, .(x = sum(covid == "negative")), by = gender]

# gender proportion tables across pos/neg 
table_gender_pos_neg <- table(d$gender, d$covid, useNA = "ifany")
margin.table(table_gender_pos_neg, 1) # calculate total genders
margin.table(table_gender_pos_neg, 2) # calculate all pos/neg  
prop.table(table_gender_pos_neg,1) # calculates proportions of variable across rows 
prop.table(table_gender_pos_neg,2) # calculates proportions of variable across columns 

# fisher exact test
fisher.test(table(d$gender, d$covid)) # without NA p = 0.496
fisher.test(table_gender_pos_neg) # with NA: p-value = 0.496
```

# fisher exact test on gender contingency table across severity 
```{r}
# severity count by gender 
covid_pos[, .(x = sum(severity == 1)), by = gender]
covid_pos[, .(x = sum(severity == 2)), by = gender]
covid_pos[, .(x = sum(severity == 3)), by = gender]
covid_pos[, .(x = sum(severity == 4)), by = gender]

# gender proportion tables across severity 
table_gender_severity <- table(covid_pos$gender, covid_pos$severity, useNA = "ifany")
table_gender_severity
margin.table(table_gender_severity, 1) # calculate total genders
margin.table(table_gender_severity, 2) # calculate all genders across severity levels 
prop.table(table_gender_severity,1) # calculates proportions of variable across rows 
prop.table(table_gender_severity,2) # calculates proportions of variable across columns 
round(100*prop.table(table_gender_severity,2), 1)

# fisher exact test
fisher.test(table(covid_pos$gender, covid_pos$severity)) # without NA p = 0.0885
fisher.test(table_gender_severity) # with NA: p-value = 0.0004022
```

# smoking status baseline statistics by severity
```{r}
# 0 - no, 1 - yes, 2 - former, 3 - unknown, NA - Unknown 
table_smoking_severity_NA <- table(covid_pos$smoking_status, covid_pos$severity, useNA = "always")
table_smoking_severity <- table(covid_pos$smoking_status, covid_pos$severity)

margin.table(table_smoking_severity_NA, 1) # calculate total smoking_status
round(100*((margin.table(table_smoking_severity_NA, 1)/91)), 1)

#margin.table(table_smoking_severity_NA, 2) # calculate all smoking_status across severity levels

#prop.table(table_smoking_severity_NA,1) # calculates proportions of variable across rows 

table_smoking_severity_NA
prop.table(table_smoking_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_smoking_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_smoking_severity) # without NA pval p = 0.644
```

# mean charleson comorbidity index baseline statistics by severity
```{r}
# summary statistics on cci by severity 
cci_severity <- covid_pos[, .(CCI), by = severity]
cci_severity[!is.na(CCI), .N, by = severity]
cci_severity[!is.na(CCI), .(x_bar = round(mean(CCI), 1), x_sd = round(sd(CCI), 1), x_range = round(range(CCI), 1)), by = severity]
# overall mean, sd, and range 
cci_severity[!is.na(CCI), .(x_bar = mean(CCI), x_sd = sd(CCI), x_range = range(CCI))]

# anova of CCI amongst severity groups 
cci_severity_model <- aov(cci_severity$CCI ~ factor(cci_severity$severity))
summary(cci_severity_model) #H0: Same mean; #H1: different means 
# p = 0.078, we can assume similar means 
```

# hospital transfered baseline statistics by severity 
```{r}
# 0 - no; # 1 - yes, NA 
table_transfer_severity_NA <- table(covid_pos$hosp_transfer, covid_pos$severity, useNA = "always")
table_transfer_severity <- table(covid_pos$hosp_transfer, covid_pos$severity)


margin.table(table_transfer_severity_NA, 1) # calculate total hosp_transfer
round(100*(margin.table(table_transfer_severity_NA, 1)/91), 1)

#margin.table(table_transfer_severity_NA, 2) # calculate total group n across severity levels
table_transfer_severity_NA # calculate each group n across severity 
#prop.table(table_transfer_severity_NA,1) # calculates proportions of variable across rows 
prop.table(table_transfer_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_transfer_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_transfer_severity) # without NA pval p = 8.157e-06 
```

# admitted to icu baseline statistics by severity 
```{r}
# 0 - no; # 1 - yes, NA 
table_icu_severity_NA <- table(covid_pos$icu_admit_UMCU, covid_pos$severity, useNA = "always")
table_icu_severity <- table(covid_pos$icu_admit_UMCU, covid_pos$severity)


margin.table(table_icu_severity_NA, 1) # calculate total icu_admit_UMCU
round(100*(margin.table(table_icu_severity_NA, 1)/189), 1)

#margin.table(table_icu_severity_NA, 2) # calculate total group n across severity levels
table_icu_severity_NA # calculate each group n across severity 
#prop.table(table_icu_severity_NA,1) # calculates proportions of variable across rows 
prop.table(table_icu_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_icu_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_icu_severity) # without NA pval p = 0.01452 
```

# pre-antibiotic use baseline statistics 
```{r}
# 1 - yes, NA - unknown
table_pre_antibiotic_severity_NA <- table(covid_pos$pre_antibiotic_use_2, covid_pos$severity, useNA = "always")
table_pre_antibiotic_severity <- table(covid_pos$pre_antibiotic_use_2, covid_pos$severity, useNA = "no")


margin.table(table_pre_antibiotic_severity_NA, 1) # calculate total pre_antibiotic_use_2
round(100*(margin.table(table_pre_antibiotic_severity_NA, 1)/91), 1)

#margin.table(table_pre_antibiotic_severity_NA, 2) # calculate total group n across severity levels
table_pre_antibiotic_severity_NA # calculate each group n across severity 
#prop.table(table_pre_antibiotic_severity_NA,1) # calculates proportions of variable across rows 
#prop.table(table_pre_antibiotic_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_pre_antibiotic_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_pre_antibiotic_severity_NA) # with NA pval p = 0.0001291
```

# mean bacterial load baseline statistics 
```{r}
# summary statistics on bacterial load by severity 
bacterial_dna_severity <- covid_pos[, .(Ct), by = severity]
bacterial_dna_severity[!is.na(Ct), .N, by = severity]

# overall mean, sd, and range
bacterial_dna_severity[!is.na(Ct), .(x_bar = round(mean(Ct),1), x_sd = round(sd(Ct), 1), x_range = round(range(Ct), 1))]

# overall mean, sd, and range per group
bacterial_dna_severity[!is.na(Ct), .(x_bar = round(mean(Ct),1), x_sd = round(sd(Ct), 1), x_range = round(range(Ct), 1)), by = severity]

# anova of Ct amongst severity groups 
bacterial_dna_severity_model <- aov(bacterial_dna_severity$Ct ~ factor(bacterial_dna_severity$severity))
summary(bacterial_dna_severity_model) #H0: Same mean; #H1: different means 
# p = 0.888, we can assume similar means 
```

# fever at admission baseline statistics
```{r}
# 0 - no, 1 - yes, NA - unknown
table_fever_severity_NA <- table(covid_pos$fever_ceoccur_v2, covid_pos$severity, useNA = "always")
table_fever_severity <- table(covid_pos$fever_ceoccur_v2, covid_pos$severity, useNA = "no")

# calculate total fever_ceoccur_v2
margin.table(table_fever_severity_NA, 1) 
round(100*(margin.table(table_fever_severity_NA, 1)/189), 1)

#margin.table(table_fever_severity_NA, 2) # calculate total group n across severity levels
table_fever_severity_NA # calculate each group n across severity 
#prop.table(table_fever_severity_NA,1) # calculates proportions of variable across rows 
#prop.table(table_fever_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_fever_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_fever_severity) # pval p = 0.01015
```

# oxygen at admisison basleine statistics 
```{r}
# 0 - no, 1 - yes, NA - unknown
table_oxygen_severity_NA <- table(covid_pos$oxygen_cmoccur, covid_pos$severity, useNA = "always")
table_oxygen_severity <- table(covid_pos$oxygen_cmoccur, covid_pos$severity, useNA = "no")


margin.table(table_oxygen_severity_NA, 1) # calculate total oxygen_cmoccur
round(100*(margin.table(table_oxygen_severity_NA, 1)/189), 1)

# margin.table(table_oxygen_severity_NA, 2) # calculate total group n across severity levels
table_oxygen_severity_NA # calculate each group n across severity 
# prop.table(table_oxygen_severity_NA,1) # calculates proportions of variable across rows 
# prop.table(table_oxygen_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_oxygen_severity_NA,2)), 1)

# fisher exact test 
fisher.test(table_oxygen_severity) # pval p = 0.2176
```

# fatigue at admission baseline statistics 
```{r}
# 0 - no, 1 - yes, NA - unknown
table_fatigue_severity_NA <- table(covid_pos$fatigue, covid_pos$severity, useNA = "always")
table_fatigue_severity <- table(covid_pos$fatigue, covid_pos$severity, useNA = "no")


margin.table(table_fatigue_severity_NA, 1) # calculate total fatigue
round(100*(margin.table(table_fatigue_severity_NA, 1)/189), 1)

#margin.table(table_fatigue_severity_NA, 2) # calculate total group n across severity levels
table_fatigue_severity_NA # calculate each group n across severity 
#prop.table(table_fatigue_severity_NA,1) # calculates proportions of variable across rows 
#prop.table(table_fatigue_severity_NA,2) # calculates proportions of variable across columns
round(100*(prop.table(table_fatigue_severity_NA,2)), 1)

# fisher exact test
fisher.test(table_fatigue_severity) # with pval p = 0.087
```

```{r}
# plots to display baseline data
ageGender <- ggplot(data = d, 
                    aes(x = gender, y = age, fill = gender)) +
  geom_boxplot() +
  xlab("gender") +
  ylab("age")
ageGender


severityPlot <- ggplot(data = d, 
                       aes(x = factor(severity), y = age, color = severity)) +
  geom_boxplot() +
  xlab("Severity Classes 1 - 4") +
  ylab("Age")
severityPlot

# deaths 
deathPlot <- ggplot(data = d, 
                    aes(x = factor(death), y = age, color = death)) +
  geom_boxplot() +
  xlab("Death") +
  ylab("Age")
deathPlot

posPlot <- ggplot(data = d, 
                  aes(x = factor(covid), y = Age, color = covid)) +
  geom_boxplot() +
  xlab("covid") +
  ylab("Age")
posPlot

ageSeverityScatter <- ggplot(data = d, 
                             aes(x = Age, y = severity_class, color = Age)) +
  geom_point()
ageSeverityScatter

countCovidAge <- ggplot(d, aes(x = Age)) + 
  geom_histogram(aes(fill = covid))
countCovidAge

countDeath <- ggplot(d, aes(x = Age)) + 
  geom_histogram(aes(fill = factor(death)))
countDeath

countQPCR <- ggplot(d, aes(x = `ng/ul`)) + 
  geom_histogram(aes(fill = `ng/ul`))
countQPCR

countGender <- ggplot(d, aes(x = Age)) + 
  geom_histogram(aes(fill = factor(gender)))
countGender
```

rm(list = ls(all = T))